\normalfont{A bird's-eye view on the habitability of exoplanets via statistical learning techniques}

### 2020-10-27 Discussion/Presentation.tex
PUB

\begin{frame}
\frametitle{Overview}
\begin{itemize}
\item\textbf{FINAL GOAL}: Survey the performances of different statistical learning algorithm in the prediction of exoplanets habitability
\item\textbf{DATASET}: Planetary Habitability Laboratory @ UPR Arecibo \cite{exoplanets-catalog}
\item\textbf{ALGORITHMS}: Decision Tree, Random Forest, Support Vector Classifier, Logistic Regression, Linear and Quadratic Classifier
\end{itemize}
\end{frame}

\section{Data}

\begin{frame}
\frametitle{Exoplanets habitability - theoretical background}
\begin{columns}
\column{0.5\textwidth}
\begin{itemize}

\item\textbf{HABITABILITY}: Rocky planets where water is present in liquid phase 
\item\textbf{LIQUID PHASE OF WATER}: At first order, if water is present, the liquid phase is controlled by the surface temperature 
\item\textbf{ATMOSPHERE}: The atmosphere (CO$_{2}$) influences the surface temperature trough the greenhouse effect
\item\textbf{H$_{2}$ and CH$_{4}$}: Other gases such as H$_{2}$ and CH$_{4}$ can produce the greenhouse effect, thus the habitable zone can be extended 
\end{itemize}
\column{0.4\textwidth}
\begin{minipage}{\textwidth}
  \centering
  \includegraphics[width=\linewidth,]{Pic/Planets_habitability_Seager.png}
\end{minipage}%

@@

>>>>

\begin{frame}
\frametitle{Overview}
\begin{itemize}
\item\textbf{Final goal}: Survey the performances of different statistical learning algorithm in the prediction of exoplanets habitability
\item\textbf{Dataset}: Planetary Habitability Laboratory @ UPR Arecibo \cite{planet_dataset}
\item\textbf{Algorithms}: Decision Tree, Random Forest, Support Vector Classifier, Logistic Regression, Linear and Quadratic Classifier
\end{itemize}
\end{frame}

\section{Theoretical background}
\subsection{Habitability}
\begin{frame}
\frametitle{Theoretical background - Exoplanets habitability }
\begin{columns}
\column{0.5\textwidth}
\begin{itemize}
\item\textbf{Habitability}: Rocky planets where water is present in liquid phase 
\item\textbf{Liquid phase}: At first order, if water is present, the liquid phase is controlled by the surface temperature 
\item\textbf{Atmosphere}: The atmosphere (CO$_{2}$) influences the surface temperature trough the greenhouse effect
\item\textbf{H$_{2}$ and CH$_{4}$}: Other gases such as H$_{2}$ and CH$_{4}$ can produce the greenhouse effect, thus the habitable zone can be extended 
\end{itemize}
\column{0.4\textwidth}
\includegraphics[width=\linewidth,]{Pic/Planets_habitability_Seager.png}\\
\begin{center}
Image taken from \cite{seager2013exoplanet}
\end{center}

@@

\subsection{Stars Feature}

\begin{frame}
\frametitle{Theoretical background - Star features }
\begin{columns}
\column{0.4\textwidth}
\begin{itemize}
\item \textbf{Main features}: For this work the main features of star are the stellar luminosity, its temperature and spectral type
\item\textbf{H-R diagram}: with these features the Hertzsprung-Russell diagram classify the stars (the temperature and spectral type of a star are two faces of the same medal)
\end{itemize}
\column{0.45\textwidth}
\includegraphics[width=\linewidth,]{Pic/S_T_T_explanation.png}
\begin{center}
Image taken from \cite{HR_diagram}
\end{center}
\end{columns}
\end{frame}

\subsection{Planet Features}

\begin{frame}
\frametitle{Theoretical background - Planet features }
\begin{center}
\includegraphics[width=.9\linewidth,]{Pic/PT_Confirmed.jpg}
\end{center}
\begin{center}
Image taken from \cite{exoplanets-catalog}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Theoretical background - Planet features }
\begin{itemize}
\item\textbf{Distance}: in this work the mean planet distance from the host star (P\_D), the periastron (P\_PN) and the apastron (P\_A) as well the termal effective distance (P\_D\_E) from the host star  were considered. These quantities constrain the planet orbital period (P\_P) via the 3$^{th}$ Kepler law (a corollary of Newton's law of universal gravitation)
\item\textbf{Mass and Radius}: the (estimated) planet mass (P\_M) and its radius (P\_R) were considered (these are also useful to distinguish the super-earth planets) 
\item\textbf{Temperature}: the planet equilibrium temperature (P\_T\_E) defined according to the expression $T_{eq}=T_{star}\sqrt{R/2a}\left(1-A\right)^{0.25}$) where R is the star radius (S\_R), a the planet mean distance (P\_D) and A the albedo here considered as 0.3 was considered as well the planet mean stellar flux $P_F$ 
\item\textbf{Habitability}: The planet habitability was classified with a boolean variable using the values reported in the dataset \cite{planet_dataset}

\end{itemize}
\end{frame}

### 2020-10-27 Manuscript/Manuscript.tex
PUB

\subsection{The decision tree and the random forest}
The first approach that was considered is the classification decision tree. First the complexity parameter (CP) was checked (a 10-fold cross-validation on the training set \footnote{In this work, for all the approaches undertaken, the training ensemble consist in a set of 350 exoplanets while the test one is composed by 150 items. } was undertaken): as can be seen from Fig \ref{Decision_tree_performances} the default parameter $0.01$ can be safely used. Then the decision tree algorithm was applied to the dataset with the tuned parameter: as it can be seen from Fig \ref{DecisionTree}, where the decision tree is given accompanied by a plot that provides the variables importance,  the main discriminant parameters, as expected, are the stellar temperature $S\_T$ and the planet temperature equilibrium $P\_T\_E$: this features discriminates, as one expects, the largest part of the habitable planet from the non-habitable. The performance of this algorithm on this dataset, evaluated with a 10-fold cross-validation \footnote{For this procedure the test set and the validation sets where reshuffled 10 times by modifying the numerical seed by which the exoplanets are assigned to the training or to the test set. The proportion 150/500 for test and training sets was keep constant}, is reported in Fig \ref{perf_table} and in the table \ref{perf_table}. Next, in order to improve the performance of this classification, the random forest was considered. As shown in Fig. \ref{Random_forest_tree_perfs} the main parameters for this algorithm were previously checked. Its results are reported in Fig \ref{Random_forest_tree_results}: here the most important variable became the $P\_T\_E$ followed by the $S\_T$: such swap is resonable since the $P\_T\_E$ determinate directly the state of water on the exoplanet. Furthermore the importance of the planet period is increased: such behaviour can be explained by considering that an habitable planet needs  to have a night-day cycle not too long in order to have a stable temperature that allows the water to keep the liquid state.  This reshuffling of the variable importance positively reflects on the performances of the model when the overall CV (as described before) is performed: as one can note from Fig the performances are slightly improved with respect to the ones reported in Fig \ref{Decision_tree_performances}, in particular the ratio of the false positive is reduced with a percentage of 40 $\%$. Such improvement is captured by the accuracy and $\phi$ factors reported in table \ref{perf_table}. Finally it is worth noting that, although  different features were also considered and the output of the response function has a larger set,  a similar hierarchy  among the variable importance for the random forest algorithm was obtained by Saha et al. \cite{saha2018machine}: in their paper the highest variable importance, via random forest algorithm,  is assigned to the mean planet temperature at surface followed by the stellar magnitude with respect to the planet (between these two there are other features strictly connected to the planet surface temperature). The last quantity is strictly connected to the stellar temperature and to the stellar spectral type. Moreover also in their study the mean distance between the planet and host star plays a minor role.

>>>>

\subsection{The decision tree and the random forest}
The first approach that was considered is the classification decision tree. First the complexity parameter (CP) was checked (a 10-fold cross-validation on the training set \footnote{In this work, for all the approaches undertaken, the training ensemble consist in a set of 350 exoplanets while the test one is composed by 150 items. } was undertaken): as can be seen from Fig \ref{Decision_tree_performances} the default parameter $0.01$ can be safely used. Then the decision tree algorithm was applied to the dataset with the tuned parameter: as it can be seen from Fig \ref{DecisionTree}, where the decision tree is given accompanied by a plot that provides the variables importance,  the main discriminant parameters, as expected, are the stellar temperature $S\_T$ and the planet temperature equilibrium $P\_T\_E$: this features discriminates, as one expects, the largest part of the habitable planet from the non-habitable. The performance of this algorithm on this dataset, evaluated with a 10-fold cross-validation \footnote{For this procedure the test set and the validation sets where reshuffled 10 times by modifying the numerical seed by which the exoplanets are assigned to the training or to the test set. The proportion 150/500 for test and training sets was keep constant}, is reported in Fig \ref{perf_table} and in the table \ref{perf_table}. Next, in order to improve the performance of this classification, the random forest was considered. As shown in Fig. \ref{Random_forest_tree_perfs} the main parameters for this algorithm were previously checked. Its results are reported in Fig \ref{Random_forest_tree_results}: here the most important variable became the $P\_T\_E$ followed by the $S\_T$: such swap is resonable since the $P\_T\_E$ determinate directly the state of water on the exoplanet. Furthermore the importance of the planet period is increased.  This reshuffling of the variable importance positively reflects on the performances of the model when the overall CV (as described before) is performed: as one can note from Fig the performances are slightly improved with respect to the ones reported in Fig \ref{Decision_tree_performances}, in particular the ratio of the false positive is reduced with a percentage of 40 $\%$. Such improvement is captured by the accuracy and $\phi$ factors reported in table \ref{perf_table}. Finally it is worth noting that, although  different features were also considered and the output of the response function has a larger set,  a similar hierarchy  among the variable importance for the random forest algorithm was obtained by Saha et al. \cite{saha2018machine}: in their paper the highest variable importance, via random forest algorithm,  is assigned to the mean planet temperature at surface followed by the stellar magnitude with respect to the planet (between these two there are other features strictly connected to the planet surface temperature). The last quantity is strictly connected to the stellar temperature and to the stellar spectral type. Moreover also in their study the mean distance between the planet and host star plays a minor role.
